A Study on Common Bug Fix Patterns in Rust

### 2022-11-02 conclusion.tex
flattened everything ;((
added Ruxanne as the name

In this project, we mined the top 18 most starred Rust projects in Github to discover common bug fix patterns lied within their code changes. We proposed a novel program modelling approach to encode most important aspects of a code change in a fixed sized datapoint. We stored 87,726 datapoints drawn from 57,214 commits across these 18 projects. After using the DBSCAN clustering algorithm, and a subsequent manual analysis, we obtained 10 clusters of general fix patterns and 9 clusters of borrow-checker related fix patterns.

>>>>

In this project, we developed Ruxanne, a tool for mining bug fix patterns in Rust. Using this tool, we mined the top 18 most starred Rust projects in Github to discover common bug fix patterns lied within their code changes. At the heart of Ruxanne, we used a novel code embedding approach to embed most important aspects of a code change in a fixed sized datapoint. We stored 87,726 datapoints drawn from 57,214 commits across these 18 projects. After using the DBSCAN clustering algorithm, and a subsequent manual analysis, we obtained 12 clusters of general fix patterns and 8 clusters of borrow-checker related fix patterns.

### 2022-11-02 introduction.tex
comment about rq1, and rephrase
added Ruxanne as the name

Producing bugs has been an unavoidable part of computer programming. Researchers have been trying to help software engineers to be more aware of the bug producing practices in development and how to step towards creating bug-free software. One useful step towards that goal would be studying common bug patterns (and their corresponding fixes) that occur across multiple software systems. Research efforts for categorizing common bug patterns date back to 1975 where Edres tried to categorize various bug patterns which occurred in operating system implementations.

In this work, we empirically analyze the most popular Rust projects in GitHub, aiming to discover common bug patterns that these projects share. Our interest comes from a desire to develop program repair tools, but we believe these insights can also be of broader service. Automated program repair tools try to locate bug locations in the source code (using a fault localization module) and then generate patches to fix them, possibly without human intervention, so that the modified program meets desired specifications (implicit or explicit). A powerful automated program repair system has many useful applications and could significantly simplify debugging, thus reducing software development cost. A set of common bug fix patterns can be a useful input to an APR system: it directly provides information about fixing similar buggy structures, which can significantly reduce the search space for patch generation.

Similarly, if we are observing undesired behaviour from our program, that would indicate that there lies a fault within our program. But where is the fault? Fault localization is the process of automatically finding the faulty statement(s) in our program. If a fault localization tool knows about the common bug patterns that can appear in the underlying programming language, it can prioritize code segment candidates based on their potentials of producing bugs. That would result in a more robust tool that leverages historical data.

Moreover, knowing which code patterns in a programming language are more susceptible to bugs can help programmers to be more cautious while working with similar code structures. This knowledge can make the developers avoid certain mistakes during production which ultimately would result in reducing the frequent bug patterns within the program. As Monperrus describes, the fewer frequent bug patterns we keep in our software, the less cost we would have to pay for software maintenance.

Moving further afield, researchers have been exploiting deep learning techniques to learn about the semantics of computer programs and create program modifier models that have a specific goal (e.g. bug fix). However, most learning tools get fixed-size vectors as input and map them to outputs. Unlike image data, it is challenging to compress programs into vectors without any semantic information loss. Current approaches try to define the vector dimensions in a way that accounts for all possible abstract syntax tree forms. That results in a large, sparse space for the inputs. Assuming that the deep learning model we are trying to develop is associated with bug fixing tasks, then knowledge about common bug fix patterns can help prune the input space towards containing fewer dimensions. That can help with creating more efficient models.

Many studies have investigated detecting and categorizing bug fix patterns in general, or for a specific programming language. In this work, we study common bug patterns that appear in Rust projects. Rust is a relatively new language that allows programmers to write programs that have low-level control over resources, as well as high-level safety guarantees. What distinguishes Rust from the other systems programming languages, e.g. C/C++, is its ownership framework, which is an integral part of Rust. The ownership framework keeps track of references to different memory locations, and will free them once they go out of scope. If a program passes the compiler’s checks, it must meet several safety properties; e.g. it has no dangling references or double-frees. These problems had to be manually ensured in languages like C/C++. Moreover, not only does the ownership concept remove the need for a garbage collector in the runtime environment, it does not require the programmer to explicitly do certain safety tasks. Due to the memory and thread safety guarantees that Rust provides, many software companies adopted Rust in production. For example, Dropbox decided to migrate to Rust for developing their storage system, since they were not completely satisfied with the features provided by AWS’s S3 or Golang.

In Rust, the so-called Borrow Checker component enforces ownership rules. Because of the Borrow Checker's importance in Rust, we split patterns into two groups: (1) non-borrow-checker related patterns (General Patterns), and (2) borrow-checker related patterns (BC-Related Patterns). We present our discovered insights of our empirical study in the following four research
questions (RQs):

\textbf{RQ1. How successful is our data modelling approach in representing program change?}

\textbf{RQ2. What are the general fix patterns in Rust, and how often do they apply?}

\textbf{RQ3. What are borrow-checker related fix patterns in Rust, and how often do they apply?}

>>>>

Bugs have long been an unavoidable part of computer programming, predating Grace Hopper's literal debugging session in 1947; even today, bug-free software remains an aspirational aim for software engineering research. One step towards that aim has been studying common bug patterns (and their corresponding fixes) across multiple software systems. Research efforts categorizing common bug patterns date back to at least 1975, where Endres tried to categorize bug patterns in operating system implementations~\cite{endres1975analysis}.

In this work, with the help of our tool, Ruxanne, we empirically analyze the most popular Rust projects in GitHub. Our goal is to discover common bug patterns that afflict these projects. Although we are motivated by a desire to develop automated program repair tools, we believe these insights can also be of broader service. Automated program repair~\cite{goues2019automated,liu2018survey} tools try to find bug locations in the source code (using a fault localization module) and then generate patches to fix them, possibly without human intervention, so that the modified program meets desired specifications (implicit or explicit). A powerful automated program repair system has many useful applications and could significantly simplify debugging, thus reducing software development cost~\cite{le2012systematic}. A set of common bug fix patterns can be a useful input to an APR system: it tells how to fix recurring buggy structures, thus reducing the search space for patch generation~\cite{jeffrey2009bugfix}.

If we observe undesired behaviour from our program, then there is likely a fault within our program\footnote{At the extreme, a fault could be caused by moths or other hardware malfunctions.}. But where is the fault? Fault localization~\cite{wong2016survey} is the process of automatically finding faulty statements in a program. If a fault localization tool knows about the common bug patterns that can appear in the underlying programming language, it can prioritize candidates based on their bug-producing potential. That would result in a more robust tool---one that leverages historical data.

Moreover, knowing that certain code patterns in a programming language are more susceptible to bugs enables projects to set policies that avoid such patterns. For instance, in the context of C and C++, Cannon et al~\cite{Cannon1997RecommendedCS} recommend that, if at least one of the if/else sections is a compound statement, requiring braces, then both sections should have braces (i.e. they should be fully bracketed). This recommendation presumably comes from qualitative experience, as it pre-dates large-scale empirical studies. A project can choose to require that contributions follow such recommendations. Research such as ours can help formulate data-driven recommendations that reduce the number of patterns empirically linked to bugs. Per Monperrus, the fewer frequent bug patterns we keep in our software, the lower cost we can expect to pay for software maintenance~\cite{monperrus2014critical}.

Moving further afield, researchers have been exploiting deep learning techniques to create program modifier models, which can then be used for specific goals (e.g. bug fixes)~\cite{alon2019code2vec,alon2018general,raychev2016learning,bielik2016phog}. However, most extant learning tools take fixed-size vectors as inputs. The process of compressing a variably-sized program to a fixed-sized vector is called code embedding~\cite{chen2019literature}. Unlike with image data, it is challenging to compress programs into vectors without any semantic information loss. Current approaches try to define the dimensions of the fixed-sized vectors in a way that accounts for all possible abstract syntax tree forms. That results in a large, sparse space for the inputs. Assuming that the deep learning model we are trying to develop is used for bug fixing, then knowledge about common bug fix patterns can help prune the dimensionality of the input space, which can help create more efficient embeddings.

Many studies have investigated detecting and categorizing bug fix patterns in general~\cite{islam2021changes,madeiral2018towards,pan2009toward}, or for a specific programming language~\cite{yang2022mining,hanam2016discovering,campos2019discovering}. In this work, we study common bug patterns that appear in Rust projects. Rust is a relatively new programming language~\cite{klabnik2019rust} which allows programmers to write programs that have low-level control over resources, while still ensuring high-level safety guarantees (for programs written in safe Rust). What distinguishes Rust from the other systems programming languages, e.g. C/C++, is its ownership framework, which is an integral part of Rust. The ownership framework keeps track of references to different memory locations, and will free them once they go out of scope. If a program passes the compiler’s checks, it meets several safety properties by construction; e.g. it has no dangling references or double-frees. Such properties had to be manually ensured (or not!) in languages like C/C++. Moreover, not only does ownership remove the need for a garbage collector in the runtime environment, it frees the programmer from the need to explicitly do certain safety-related tasks~\cite{qin2020understanding}. Due to the memory and thread safety guarantees that Rust provides, many software companies have adopted Rust in production. For example, Dropbox decided to migrate to Rust for developing their storage system, since they were not completely satisfied with the features provided by AWS’s S3 or Golang~\cite{dropbox}.

The Rust compiler's Borrow-Checker component enforces ownership rules and hence many of Rust's safety properties. Because of the Borrow-Checker's importance in Rust, we split the patterns that we present into two groups: (1) non-borrow-checker related patterns (General Patterns), and (2) borrow-checker related patterns (BC-Related Patterns). \\

\noindent \textbf{Research Questions} We formulated three research questions (RQs) which guided our empirical study about Rust bug patterns:

\begin{itemize}
    \item \textbf{RQ1. Does our data modelling approach capture the most important aspects of bug-fixing program changes?} 
    \item \textbf{RQ2. What are the general fix patterns in Rust, and how often do they apply?} 
    \item \textbf{RQ3. What are borrow-checker related fix patterns in Rust, and how often do they apply?}
\end{itemize}

\noindent \textbf{Contributions} This paper makes the following primary contributions:

\begin{itemize}
    \item To the best of our knowledge, this work is the first to exploits a code analysis pipeline to automatically mine Rust open source projects and extract common cross-project bug fix patterns.
    \item We propose a novel method to encode key AST information in fixed size datapoints using a semi-automatically derived weighting scheme.
    \item To respect the key role of the borrow checker in Rust, we report and discuss bug fix patterns that are related to the ownership framework in Rust separately from general patterns.
\end{itemize}

### 2022-11-02 main.tex
shuffle package order
flattened everything ;((
methodology comments; probably need to move things around
added Ruxanne as the name

\usepackage{listings, listings-rust}
\lstset{escapeinside={<@}{@>}}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{array}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage[utf8]{inputenc}
\usepackage{tikz}
\usetikzlibrary{graphs} 
\usetikzlibrary{matrix}
\usepackage{multirow}
% \usepackage[backend=bibtex, style=authoryear]{biblatex}
% \addbibresource{bibtex.bib}
\usepackage{subfiles} % Best loaded last in the preamble
\input{defs}

@@

Yet, to the best of our knowledge, no research has sought to identify the most pervasive bug fix patterns within Rust programs. In this project, we developed a novel embedding of Rust code into fixed-sized vectors. Using this embedding, we mined the top 18 most-starred Rust projects in Github to discover the most common bug fix patterns committed to their repositories. We analyzed 87,726 code changes drawn from 57,214 commits across these 18 projects. After clustering the code changes, and conducting a manual analysis, we discovered 19 groups of cross-project bug fix patterns, which we categorize as (1) general patterns and (2) borrow-checker-related patterns. We describe each of these patterns and their implications to automated program repair.
\keywords{First keyword \and Second keyword \and More}
% \PACS{PACS code1 \and PACS code2 \and more}
\subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
% rubber: depend sections/introduction.tex
\subfile{sections/introduction}

\section{Methodology}
% rubber: depend sections/methodology.tex
\subfile{sections/methodology}

\section{Results}
% rubber: depend sections/results.tex
\subfile{sections/results}

\section{Discussion}
% rubber: depend sections/discussion.tex
\subfile{sections/discussion}

\section{Related Work}
% rubber: depend sections/related.tex
\subfile{sections/related}

\section{Conclusion}
% rubber: depend sections/conclusion.tex
\subfile{sections/conclusion}

@@

\section{Appendix}
\subfile{sections/appendix}

>>>>

\usepackage{listings, listings-rust}
\lstset{escapeinside={<@}{@>}}
\usepackage{tabularx}
\usepackage{array}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage[utf8]{inputenc}
\usepackage{tikz}
\usetikzlibrary{graphs} 
\usetikzlibrary{matrix}
\usepackage{multirow}
% \usepackage[backend=bibtex, style=authoryear]{biblatex}
% \addbibresource{bibtex.bib}
\usepackage{todonotes}
\usepackage{hyperref}
\usepackage{subfiles} % Best loaded last in the preamble
\input{defs}

@@

Yet, to the best of our knowledge, no research has sought to identify the most pervasive bug fix patterns within Rust programs. In this project, we introduce Ruxanne, a tool for analyzing and extracting fix patterns in Rust. Ruxanne uses a novel embedding of Rust code into fixed-sized vectors. Through Ruxanne, we mined the top 18 most-starred Rust projects in Github to discover the most common bug fix patterns committed to their repositories. We analyzed 87,726 code changes drawn from 57,214 commits across these 18 projects. After clustering the code changes, and conducting a manual analysis, we discovered 20 groups of cross-project bug fix patterns, which we categorize as (1) general patterns and (2) borrow-checker-related patterns. We describe each of these patterns and their implications to automated program repair.
\keywords{First keyword \and Second keyword \and More}
% \PACS{PACS code1 \and PACS code2 \and more}
\subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
% rubber: depend introduction.tex
\subfile{introduction}

\section{Methodology}
% rubber: depend methodology.tex
\subfile{methodology}

\section{Results}
% rubber: depend results.tex
\subfile{results}

\section{Discussion}
% rubber: depend discussion.tex
\subfile{discussion}

\section{Related Work}
% rubber: depend related.tex
\subfile{related}

\section{Conclusion}
% rubber: depend conclusion.tex
\subfile{conclusion}

@@

\section{Appendix}
\subfile{appendix}

### 2022-11-02 methodology.tex
fixed discussion, results todos, added two G patterns, dropped one BC pattern
added Ruxanne as the name

In this section, we describe our methodology, providing full details in later subsections. Our pipeline for finding classes of pervasive patterns in Rust programs parses program versions, computes differences between them, embeds them in fixed-size datapoints, and then clusters these datapoints. After conducting a manual analysis on the obtained clusters, we refine them towards manifesting bug fix patterns.

@@

We decided to implement our pipeline completely in Python. Our target repositories were the top 18 most-starred open source Rust projects on GitHub as of August 2021. We mined (using Pydriller) all the bug related commits and ran them through our pipeline; Section~\ref{sec:data_modelling} describes how we extracted fixed-size datapoints from the code changes. Section~\ref{sec:mining} describes our mining methodology in depth. Then, we applied the DBSCAN clustering algorithm~\cite{ester1996density} on the resulting datapoints. Section~\ref{sec:clustering_data} describes clustering in detail, justifies our choice of DBSCAN for clustering, and explains how we tuned it to improve cluster quality.

@@

    \draw[->]  (c1) -- (d1);

@@

The second part identifies the context of the modification, that is, the path from the root of the tree to the subtree in which the modification has occured. The third part is the content of the change---terminals and nonterminals. Our tool\todo{do we have a name? It's important to have a name} looks for changes within each root-level scope (Rust book calls these root-level scopes the items of a crate\footnote{\url{https://doc.rust-lang.org/reference/items.html}}). That is, our assumption is that each change pattern occurs within one root-level scope. In this work, we chose to not handle patterns that involve changes in multiple functions or multiple files. However, our tool detects both continuous and non-continuous line changes within one scope.

>>>>

In this section, we describe our methodology, providing full details in later subsections. Ruxanne uses a pipeline for finding classes of pervasive patterns in Rust programs which parses program versions, computes differences between them, embeds them in fixed-size datapoints, and then clusters these datapoints. After conducting a manual analysis on the obtained clusters, we refine them towards manifesting bug fix patterns.

@@

We decided to implement Ruxanne completely in Python. Our target repositories were the top 18 most-starred open source Rust projects on GitHub as of August 2021. We mined (using Pydriller) all the bug related commits and ran them through our pipeline; Section~\ref{sec:data_modelling} describes how we extracted fixed-size datapoints from the code changes. Section~\ref{sec:mining} describes our mining methodology in depth. Then, we applied the DBSCAN clustering algorithm~\cite{ester1996density} on the resulting datapoints. Section~\ref{sec:clustering_data} describes clustering in detail, justifies our choice of DBSCAN for clustering, and explains how we tuned it to improve cluster quality.

@@

    \draw[->]  (c1) -- (d1);
    \draw[->]  (d1) -- (d2);

@@

The second part identifies the context of the modification, that is, the path from the root of the tree to the subtree in which the modification has occured. The third part is the content of the change---terminals and nonterminals. Ruxanne looks for changes within each root-level scope (Rust book calls these root-level scopes the items of a crate\footnote{\url{https://doc.rust-lang.org/reference/items.html}}). That is, our assumption is that each change pattern occurs within one root-level scope. In this work, we chose to not handle patterns that involve changes in multiple functions or multiple files. However, our tool detects both continuous and non-continuous line changes within one scope.
